Structural Breaks in Inflation Dynamics within the European Monetary Union

### 2011-05-15 pkg/paper/Paper/draft.Rnw
updated greyplot graphics
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/glogis@92 b679b2f7-6539-45d0-a605-5ddcb976cde9

## include glmean dataset
if(!file.exists(file.path(root_dir, "Analysis", "glmean.rda"))) {
  glmean <- function(x) {
    if(any(is.na(x))) NA else glogisfit(x, hessian = FALSE)$moments[1]
  }
  ## returns
  hicp <- diff(log(HICP[,]))
  ## rolling GL means

  hicp_glmean <- rollapply(hicp, 12, glmean)

  hicp_euro <- diff(log(HICP[, c(1,2,6,7,8,9,11,12,13,14,16,19)]))
	mean_euro <- hicp_glmean[, c(1,2,6,7,8,9,11,12,13,14,16,19)]

	diff <- rep(1,length(mean_euro[,1]))
	n <- length(mean_euro[,1])

	for (i in 1:n) {
	bla <- subset(mean_euro[i,], select=colMeans(is.na(mean_euro[i,])) == 0) 
	diff[i] = max(bla) - min(bla)
	}
	diff <- ts(diff , start = c(1990,2), frequency =12)

  setwd(file.path(root_dir, "Analysis"))
	save(hicp, hicp_glmean, diff, file = "glmean.rda")
	setwd(file.path(root_dir, "Paper"))

@@

\clearpage
\section{Data}\label{s2}
%(... first method, than data...) \\
We \fixme{Jesus: yearly inflation rates better; don't start with data in paper} employ seasonally unadjusted HICP series of 21 countries ranging from January 1990 to December 2010 that are obtained from the \citet{oecd}. Countries included are Austria,	Belgium, Czech Republic,	Denmark, Estonia,	Finland,	France,	Germany,	Greece,	Hungary,	Ireland,	Italy,	Luxembourg,	Netherlands,	Poland,	Portugal,	Slovakia, Slovenia,  Spain,	Sweden, and	United Kingdom. Latvia and Lithuania as well as Bulgaria and Romania are excluded due to data scarcity. Cypria and Malta are not included since they are very small economies. \\
The countries in this sample can be divided into three different groups. First the euro countries (Austria, Belgium, Estonia, France, Finland, Germany, Greece, Ireland, Italy, Luxembourg, Netherlands, Portugal, Slovakia, Slovenia and Spain).  Second, the EU members not participating in the ERM~II: Czech Republic, Hungary, Poland, United Kingdom, and Sweden. Denmark stands on its own as a member of the EU and the ERM~II, but not yet a member of the EMU. \\
%(... schlussverkäufe schon hier erwähnen, einfach cool, ja das gibts, wir haben die aber weggeschätzt; generell viel angelsächsischer/präpotenter schreiben; nur 1 weg der richtige und zwar unserer; nicht erwähnen, dass andere dinge auch noch möglich; so nach motto: mich interessierts, ich machs, andere erst am ende erwähnen...)
% greece: -for winter discounts: First 10 days of January-end of February; -for summer discounts: First 10 days of July -end od August; but has always been included
As \readme{Jesus: Italy and Spain are problematic, fixed with same method as Belgium and Luxembourg} shown in a paper by \cite{pltm}, the harmonization of the treatment of sales prices led to considerable changes in some countries (Belgium, Italy, Luxembourg and Spain); failing to take this into account may lead to considerable distortions in the estimation. In these cases, we dealt with the inclusion of the sales periods effects\footnote{This was demanded by a Commission regulation, details may be found in: \citet{eurlex} and in the Appendix \ref{dataprobs}.} into the HICP series (January 2000 for Belgium, January 2001 for Italy, January 1999 for Luxembourg and January 2001 for Spain) by estimating dummies for these month\footnote{There are two month with a significant decrease in prices (January and July), followed by significant price increases in the following months (February and August).} and subtracting them from the original time series.

@@

The disparity of the national inflation rates is still an issue. Many researchers, like \citet{hofmann}  find a considerable amount of inflation differentials. \citet{cap} estimate short-run and steady-state inflation uncertainty in 12 EMU countries and find a considerable degree of heterogeneity across EMU countries in terms of average inflation and its degree of persistence. In a paper examining structural convergence of inflation rates in EU countries, \citet{palomba} try to answer the question if during the 1990s the inflation rate dynamics of EU countries become more similar. They find that convergence over time of inflation dynamics was only partly observable. 
In a paper studying core inflation and using an aggregated euro area inflation rate, \citet{morana} finds three regimes (roughly 1980---1984, 1984---1993 and 1993---2000) governing the core inflation rate.  \\
Concerning inflation differentials, Figure \ref{greyplot} depicts the inflation rates and a rolling mean estimation, where the mean is calculated using the first moment of a fitted GL distribution.\footnote{If more countries have the same inflation rate in a given month, the lines are overlapped and thus get darker. We wold see one black line if all countries would show the same month--on--month increase at any one time. The red lines depict the means.} We do see that many of these countries follow a rather similar path. One event clearly visible is the financial crises and the short deflationary phase starting somewhere around the autumn of 2007 and lasting a few months. Over time, there are hints, that the inflation rates converged towards some common -- albeit broad -- mean level. For the inflation series of the whole EMU, as published by the \citet{oecd}, we see \readme{Jesus: why not do this for EMU inflation (i.e. the one time series of the OECD} an increase in mean from \Sexpr{round(euhim1, 5)} (1990(1)--1998(12)) to \Sexpr{round(euhim2, 5)} (1999(1)--2008(8)) with a strong decrease to \Sexpr{round(euhim3, 5)} (2008(9)--2010(12)) following the financial crises. Whereas variance has increased over time from    \Sexpr{round(euhiv1, 5)} to \Sexpr{round(euhim3, 5)}. Regarding skewness, we see a clear change towards a negative skewness from \Sexpr{round(euhis1, 5)} to \Sexpr{round(euhis3, 5)}.

%(... compare this with the emu inflation rate...)\\

\begin{figure}[h!tp]
<<greyplot, fig=TRUE, height=8, width=8>>=
  gray_red <- rgb(c(0.2, 0.8), c(0.2, 0), c(0.2, 0), alpha = 0.3)
  ## plot returns (gray) and rolling means (red)
  plot(hicp, plot.type = "single", lwd = 1.5, col = gray_red[1], xlab="Time", ylab="Series")
  for(i in 1:ncol(hicp)) lines(hicp_glmean[,i], lwd = 1.5, col = gray_red[2])
@
\caption{Inflation rates and estimated GL mean over time}
\label{greyplot}
\end{figure}

@@

Regarding the general trend, Table \ref{grouptable} presents lower, medium and upper (25, 50, 75) percentiles of the first three moments of the inflation series  at the beginning of the convergence phase (January 1995), the beginning of the financial crisis (in September 2008\footnote{After the insolvency of the Lehman Brothers in September 2008.}) and for the last observation in the sample. The countries are grouped into the ``core'' of the eurozone (Austria, Belgium, Finland, France, Germany, Luxembourg and Netherlands), the PIIGS countries (Portugal, Italy, Ireland, Greece and Spain) and all the others.  \\

\begin{longtable}{|l|l|l|l|l|}
\hline
\multicolumn{5}{|c|}{1995} \\
\hline
Zone & Percentile & Mean & Variance & Skewness\\
\hline
<<echo=FALSE, results=tex>>=
euro <- c(1, 2, 6, 7, 8, 13, 14)
noneuro <- c(3, 4, 5, 10, 15, 17, 18, 20, 21)
piigs <- c(9, 11, 12, 16, 19)
writeLines(format_cross_country(cross_country_moments(euro, 1995), name ="Euro"))
@
\hline
<<echo=FALSE, results=tex>>=
writeLines(format_cross_country(cross_country_moments(piigs, 1995), name ="PIIGS"))
@
\hline
<<echo=FALSE, results=tex>>=
writeLines(format_cross_country(cross_country_moments(noneuro, 1995), name ="Noneuro"))
@
\hline
\multicolumn{5}{|c|}{2008(9)} \\
\hline
<<echo=FALSE, results=tex>>=
writeLines(format_cross_country(cross_country_moments(euro, 2008+8/12), name ="Euro"))
@
\hline
<<echo=FALSE, results=tex>>=
writeLines(format_cross_country(cross_country_moments(piigs, 2008+8/12), name ="PIIGS"))
@
\hline
<<echo=FALSE, results=tex>>=
writeLines(format_cross_country(cross_country_moments(noneuro, 2008+8/12), name ="Noneuro"))
@
\hline
\multicolumn{5}{|c|}{2010(12)} \\
\hline
<<echo=FALSE, results=tex>>=
writeLines(format_cross_country(cross_country_moments(euro, 2010+11/12), name ="Euro"))
@
\hline
<<echo=FALSE, results=tex>>=
writeLines(format_cross_country(cross_country_moments(piigs, 2010+11/12), name ="PIIGS"))
@
\hline
<<echo=FALSE, results=tex>>=
writeLines(format_cross_country(cross_country_moments(noneuro, 2010+11/12), name ="Noneuro"))
@
\hline
\caption{Development of first three moments}
\label{grouptable}
\end{longtable}

>>>>

if(!file.exists(file.path(root_dir, "Analysis", "glmean.rda"))) {
   glmean <- function(x) {
    if(any(is.na(x))) NA else glogisfit(x, hessian = FALSE)$moments[1]
  }

  hicps <- hicps[, colnames(hicps) != "EU"]

  ## rolling GL means
  hicp_glmean <- rollapply(hicps, 12, glmean)

  setwd(file.path(root_dir, "Analysis"))
	save(hicps, hicp_glmean, file = "glmean.rda")
	setwd(file.path(root_dir, "Paper"))

@@

\clearpage
\section{Data}\label{s2}
%(... first method, than data...) \\

% wichtig: http://sdw.ecb.europa.eu/quickview.do?SERIES_KEY=122.ICP.M.U2.N.000000.4.ANR
% sieht man die daten von sales prices: 1999-1 portugal,  2000-1 lux, 2001-1 bel, 2002 ita-spa (allerdings in anderem paper ein jahr früher) hm

We \fixme{Jesus: yearly inflation rates better; don't start with data in paper} employ seasonally unadjusted HICP series of 21 countries ranging from January 1990 to December 2010 that are obtained from the \citet{oecd}. Countries included are Austria,	Belgium, Czech Republic,	Denmark, Estonia,	Finland,	France,	Germany,	Greece,	Hungary,	Ireland,	Italy,	Luxembourg,	Netherlands,	Poland,	Portugal,	Slovakia, Slovenia,  Spain,	Sweden, and	United Kingdom. Latvia and Lithuania as well as Bulgaria and Romania are excluded due to data scarcity. Cypria and Malta are not included since they are very small economies. \\
The countries in this sample can be divided into three different groups. First the euro countries (Austria, Belgium, Estonia, France, Finland, Germany, Greece, Ireland, Italy, Luxembourg, Netherlands, Portugal, Slovakia, Slovenia and Spain).  Second, the EU members not participating in the ERM~II: Czech Republic, Hungary, Poland, United Kingdom, and Sweden. Denmark stands on its own as a member of the EU and the ERM~II, but not yet a member of the EMU. \\
%(... schlussverkäufe schon hier erwähnen, einfach cool, ja das gibts, wir haben die aber weggeschätzt; generell viel angelsächsischer/präpotenter schreiben; nur 1 weg der richtige und zwar unserer; nicht erwähnen, dass andere dinge auch noch möglich; so nach motto: mich interessierts, ich machs, andere erst am ende erwähnen...)
% greece: -for winter discounts: First 10 days of January-end of February; -for summer discounts: First 10 days of July -end od August; but has always been included
As \readme{Jesus: Italy and Spain are problematic, fixed with same method as Belgium and Luxembourg} shown in a paper by \cite{pltm}, the harmonization of the treatment of sales prices led to considerable changes in some countries (Belgium, Italy, Luxembourg and Spain); failing to take this into account may lead to considerable distortions in the estimation. In these cases, we dealt with the inclusion of the sales periods effects\footnote{This was demanded by a Commission regulation, details may be found in: \citet{eurlex} and in the Appendix \ref{dataprobs}.} into the HICP series (January 2000 for Belgium, January 2001 for Italy, January 1999 for Luxembourg and January 2001 for Spain) by estimating dummies for these month\footnote{There are two month with a significant decrease in prices (January and July), followed by significant price increases in the following months (February and August).} and subtracting them from the original time series.
\fixme{update for using of X-12-ARIMA method, argue with seasonal effects in \cite{pltm}: In Germany and Portugal, sales prices are included from 1998 onwards, in Ireland and Luxembourg from 1999 onwards and in Belgium from 2000. In January 2001, Italy and Spain were the last two EU15 countries to include sales into their HICP statistics (The official date may be 12 months later. However, the indices of the previous 12 months had been revised in order to avoid distortions in the inflation rates.)}

@@

The disparity of the national inflation rates is still an issue. Many researchers, like \citet{hofmann}  find a considerable amount of inflation differentials. \citet{cap} estimate short-run and steady-state inflation uncertainty in 12 EMU countries and find a considerable degree of heterogeneity across EMU countries in terms of average inflation and its degree of persistence. In a paper examining structural convergence of inflation rates in EU countries, \citet{palomba} try to answer the question if during the 1990s the inflation rate dynamics of EU countries become more similar. They find that convergence over time of inflation dynamics was only partly observable. 
In a paper studying core inflation and using an aggregated euro area inflation rate, \citet{morana} finds three regimes (roughly 1980---1984, 1984---1993 and 1993---2000) governing the core inflation rate.  \\
Concerning inflation differentials, Figure \ref{greyplot} depicts the inflation rates and a rolling mean estimation (with a window size of 12 months), where the mean is calculated using the first moment of a fitted GL distribution.\footnote{If more countries have the same inflation rate in a given month, the lines are overlapped and thus get darker. We wold see one black line if all countries would show the same month--on--month increase at any one time. The red lines depict the means.} We do see that many of these countries follow a rather similar path. One event clearly visible is the financial crises and the short deflationary phase starting somewhere around the autumn of 2007 and lasting a few months. Over time, there are hints, that the inflation rates converged towards some common -- albeit broad -- mean level. For the inflation series of the whole EMU, as published by the \citet{oecd}, we see \readme{Jesus: why not do this for EMU inflation (i.e. the one time series of the OECD} an increase in mean from \Sexpr{round(euhim1, 5)} (1990(1)--1998(12)) to \Sexpr{round(euhim2, 5)} (1999(1)--2008(8)) with a strong decrease to \Sexpr{round(euhim3, 5)} (2008(9)--2010(12)) following the financial crises. Whereas variance has increased over time from    \Sexpr{round(euhiv1, 5)} to \Sexpr{round(euhim3, 5)}. Regarding skewness, we see a clear change towards a negative skewness from \Sexpr{round(euhis1, 5)} to \Sexpr{round(euhis3, 5)}.

% if(!file.exists(file.path(root_dir, "Paper", "draft-greyplot.pdf")))

%\begin{figure}[h!tp]
%<<greyplot, fig=TRUE, height=8, width=8>>=
%
% ## compute GL mean (if no NA)
%  glmean <- function(x) {
%    if(any(is.na(x))) NA else glogisfit(x, hessian = FALSE)$moments[1]
%  }
%  hicps <- hicps[, colnames(hicps) != "EU"]
%
%  ## rolling GL means
%  hicp_glmean <- rollapply(hicps, 12, glmean)
%
%  ## semi-transparent gray and red
%  gray_red <- rgb(c(0.2, 0.8), c(0.2, 0), c(0.2, 0), alpha = 0.3)
%
%  ## plot returns (gray) and rolling means (red)
%  plot(hicps, plot.type = "single", lwd = 1.5, col = gray_red[1], xlab="Time", ylab="Series")
%  for(i in 1:ncol(hicp)) lines(hicp_glmean[,i], lwd = 1.5, col = gray_red[2])
%@
%\caption{Inflation rates and estimated GL mean over time}
%\label{greyplot}
%\end{figure}

\begin{figure}[h!tp]
<<greyplot, fig=TRUE, height=8, width=8>>=
  gray_red <- rgb(c(0.2, 0.8), c(0.2, 0), c(0.2, 0), alpha = 0.3)
  ## plot returns (gray) and rolling means (red)
  hicps <- hicps[, colnames(hicps) != "EU"]
  plot(hicps, plot.type = "single", lwd = 1.5, col = gray_red[1], xlab="Time", ylab="Series")
  for(i in 1:ncol(hicps)) lines(hicp_glmean[,i], lwd = 1.5, col = gray_red[2])
@
\caption{Inflation rates and estimated GL mean over time}
\label{greyplot}
\end{figure}

@@

Regarding the general trend, Table \ref{grouptable} presents lower, medium and upper (25, 50, 75) percentiles of the first three moments of the inflation series  at the beginning of the convergence phase (January 1995), in December 2007 as a point in time before the financial crisis  and for the last observation in the sample. The countries are grouped into the ``core'' of the eurozone (Austria, Belgium, Finland, France, Germany, Luxembourg and Netherlands), the PIIGS countries (Portugal, Italy, Ireland, Greece and Spain) and all the others.  \\

\begin{longtable}{|l|l|l|l|l|}
\hline
\multicolumn{5}{|c|}{1995} \\
\hline
Zone & Percentile & Mean & Variance & Skewness\\
\hline
<<echo=FALSE, results=tex>>=
euro <- c(1, 2, 6, 7, 8, 13, 14)
noneuro <- c(3, 4, 5, 10, 15, 17, 18, 20, 21)
piigs <- c(9, 11, 12, 16, 19)
writeLines(format_cross_country(cross_country_moments(euro, 1995), name ="Euro"))
@
\hline
<<echo=FALSE, results=tex>>=
writeLines(format_cross_country(cross_country_moments(piigs, 1995), name ="PIIGS"))
@
\hline
<<echo=FALSE, results=tex>>=
writeLines(format_cross_country(cross_country_moments(noneuro, 1995), name ="Noneuro"))
@
\hline
\multicolumn{5}{|c|}{2007(12)} \\
\hline
<<echo=FALSE, results=tex>>=
writeLines(format_cross_country(cross_country_moments(euro, 2007+11/12), name ="Euro"))
@
\hline
<<echo=FALSE, results=tex>>=
writeLines(format_cross_country(cross_country_moments(piigs, 2007+11/12), name ="PIIGS"))
@
\hline
<<echo=FALSE, results=tex>>=
writeLines(format_cross_country(cross_country_moments(noneuro, 2007+11/12), name ="Noneuro"))
@
\hline
\multicolumn{5}{|c|}{2010(12)} \\
\hline
<<echo=FALSE, results=tex>>=
writeLines(format_cross_country(cross_country_moments(euro, 2010+11/12), name ="Euro"))
@
\hline
<<echo=FALSE, results=tex>>=
writeLines(format_cross_country(cross_country_moments(piigs, 2010+11/12), name ="PIIGS"))
@
\hline
<<echo=FALSE, results=tex>>=
writeLines(format_cross_country(cross_country_moments(noneuro, 2010+11/12), name ="Noneuro"))
@
\hline
\caption{Development of first three moments}
\label{grouptable}
\end{longtable}
